---
title: "Assignment 10: Data Scraping"
author: "Lauren Ng"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse);library(rvest)
library(dplyr)
library(ggplot2)
library(lubridate)


```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2023 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2023>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
webpage <- read_html(
  'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2023')
webpage
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
watersystemname <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
watersystemname

PWSID <- webpage %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()
PWSID

ownership <- webpage %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()
ownership

maxdayuse <- webpage %>%
  html_nodes("th~ td+ td") %>%
  html_text()
maxdayuse

month <- webpage %>%
  html_nodes(".fancy-table:nth-child(31) tr+ tr th")%>%
html_text()
month
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2023, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 

df_withdrawals <- data.frame(
  "Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
  "Year" = rep(2023, 12),
  "Water System" = watersystemname,
  "PWSID" = PWSID,
  "Ownership" = ownership,
  "Max_Daily_Use" = as.numeric(maxdayuse)
) %>%
mutate(Date = make_date(Year, Month, 1))

#5 
ggplot(df_withdrawals, aes(x = Date, y = `Max_Daily_Use`)) + 
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = paste("2023 Water Usage Data for", watersystemname),
    y = "Withdrawal (mgd)",
    x = "Month"
  ) +
  theme_minimal()
```

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data, returning a dataframe.
**Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}

#6
scrape.it <- function(pwsid,the_year){

the_scrape_url <-  read_html(
paste0(
'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', pwsid,'&', 'year=',the_year)
)


#Set the element address variables (determined in the previous step)

watersystemname_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
PWSID_tag <- 'td tr:nth-child(1) td:nth-child(5)'
ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
maxdayuse_tag <- 'th~ td+ td'

#Scrape the data items
watersystemname <- the_scrape_url %>% 
html_nodes(watersystemname_tag) %>% 
html_text()
PWSID <- the_scrape_url %>%   
html_nodes(PWSID_tag) %>%  
html_text()
ownership <- the_scrape_url %>% 
html_nodes(ownership_tag) %>% 
html_text()
maxdayuse <- the_scrape_url %>%
html_nodes(maxdayuse_tag) %>% 
html_text()

#Convert to a dataframe
df_withdrawals <- data.frame(
  "Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
  "Year" = rep(the_year, 12),
  "Water System" = watersystemname,
  "PWSID" = PWSID,
  "Ownership" = ownership,
  "Max_Daily_Use" = as.numeric(maxdayuse)
) %>%
mutate(Date = make_date(Year, Month, 1))

return(df_withdrawals)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
Durham_2015 <- scrape.it('03-32-010',2015)
view(Durham_2015)

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
Asheville_2015 <- scrape.it('01-11-010',2015)
view(Asheville_2015)

combined_data <- bind_rows(
  Durham_2015 %>% mutate(Location = "Durham"),
  Asheville_2015 %>% mutate(Location = "Asheville")
)

ggplot(combined_data, aes(x = Date, y = Max_Daily_Use, color = Location)) +
  geom_line(size = 1) +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed") +
  labs(
    title = "Comparison of Max Daily Water Withdrawals (2015)",
    y = "Withdrawal (mgd)",
    x = "Month",
    color = "Location"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2018 thru 2022.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
asheville_pwsid <- "01-11-010"
years <- 2018:2022

# Scrape data for Asheville for the specified years using map2()
#asheville_data <- map2(asheville_pwsid,years, scrape.it)%>%
asheville_data <- map2_dfr(rep(asheville_pwsid, length(years)), years, scrape.it)
bind_rows(asheville_data)

ggplot(asheville_data, aes(x = Date, y = Max_Daily_Use)) +
  geom_line(size = 1) +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed") +
  labs(
    title = "Comparison of Max Daily Water Withdrawals (2015)",
    y = "Withdrawal (mgd)",
    x = "Month"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:Yes, looking at the plot, it appears that Asheville does have a trend in increasing water usage over time.
>
